---
title: "cranwatch"
subtitle: "Health radar for the R packages I use, or am eyeing, at work"
---

```{r}
#| include: false
for (f in list.files("../R", pattern = "\\.R$", full.names = TRUE)) source(f)
library(reactable)
library(htmltools)

snap  <- read_snapshot("../data/latest.csv")
run   <- read_json("../data/run.json")
links <- read_yaml("../links.yml")
repo  <- Sys.getenv("GITHUB_REPOSITORY", "cgwweaver/cranwatch")

failed    <- run$sources |> keep(\(s) !isTRUE(s$ok)) |> names()
sources   <- if (length(failed)) paste("**FAILED:**", paste(failed, collapse = ", ")) else "all ok"
note      <- \(l) if (is.null(l$note)) "" else paste0(": ", l$note)
```

**`r nrow(snap)` packages** · run `r run$run_date` · downloads `r run$dl_windows$dl_365[[1]]` to `r run$dl_windows$dl_365[[2]]` · sources `r sources` ·
[CSV](https://github.com/`r repo`/blob/main/data/latest.csv) · [history](https://github.com/`r repo`/tree/main/data/snapshots) · [edit watchlist](https://github.com/`r repo`/edit/main/packages.yml)

```{r}
#| results: asis
alerts <- alerts_of(snap) |> left_join(snap |> select(package, link), by = "package")
if (nrow(alerts)) {
  cat("::: {.callout-warning}\n## Alerts\n\n")
  cat(glue("- **[{alerts$package}]({alerts$link})**: {alerts$detail}"), sep = "\n")
  cat("\n:::\n")
}
```

```{r}
#| column: screen-inset
notes <- read_watch_notes("../packages.yml") |>
  filter(!is.na(package)) |>
  summarise(notes = paste(unique(note), collapse = "; "), .by = package)

tbl <- snap |>
  left_join(notes, by = "package") |>
  select(
    package, title, link, themes, status,
    ru_score, dl_365, dl_growth, ru_stars, ru_dependents,
    deadline, archived_n, archived_last, cran_comment,
    needs_compilation, deps_n, deps_compiled_n, sysreqs, sysreqs_via_deps, sysdeps_detected,
    revdeps_n, revdeps_top, flags, notes
  )

hidden <- c("title", "link", "archived_last", "cran_comment", "sysreqs_via_deps", "sysdeps_detected")

flag_badges <- \(value) {
  if (is.na(value)) return("")
  str_split_1(value, "; ") |> map(\(f) span(class = paste0("flag flag-", f), f)) |> tagList()
}

muted <- \(...) div(class = "muted", ...)

# Hover text for the score: what it's made of (see "Columns & sources")
score_parts <- ru_score_parts(snap, "ru_") |>
  pmap_chr(\(...) { p <- round(c(...), 1); paste0("= 1 + ", paste(p[p > 0], names(p)[p > 0], collapse = " + ")) })
score_tip <- if_else(is.na(snap$ru_update_weeks), "", score_parts)

reactable(
  tbl,
  searchable = TRUE, filterable = TRUE, highlight = TRUE, striped = TRUE, compact = TRUE,
  defaultSorted = "ru_score", defaultSortOrder = "desc",
  defaultPageSize = 50, showPageSizeOptions = TRUE, pageSizeOptions = c(25, 50, 200),
  defaultColDef = colDef(minWidth = 75, headerVAlign = "bottom", headerStyle = list(whiteSpace = "nowrap")),
  columnGroups = list(
    colGroup("Popularity", c("ru_score", "dl_365", "dl_growth", "ru_stars", "ru_dependents")),
    colGroup("CRAN health", c("deadline", "archived_n")),
    colGroup("Install burden", c("needs_compilation", "deps_n", "deps_compiled_n", "sysreqs")),
    colGroup("Reverse deps", c("revdeps_n", "revdeps_top"))
  ),
  columns = c(
    set_names(map(hidden, \(x) colDef(show = FALSE)), hidden),
    list(
      package = colDef(name = "Package", sticky = "left", minWidth = 130, cell = \(value, i) {
        # NA attributes crash reactable's React renderer, hence coalesce()
        if (is.na(tbl$link[i])) value else a(href = tbl$link[i], title = coalesce(tbl$title[i], ""), target = "_blank", value)
      }),
      themes  = colDef(name = "Themes", minWidth = 120),
      status  = colDef(name = "Status", minWidth = 100, cell = \(value) span(class = paste0("status status-", value), value)),
      ru_score      = colDef(name = "Score", cell = \(value, i) {
        if (is.na(value)) "" else span(title = score_tip[i], class = if (nzchar(score_tip[i])) "has-tip", format(round(value, 1), nsmall = 1))
      }),
      dl_365        = colDef(name = "DLs/yr", cell = fmt_short),
      dl_growth     = colDef(name = "vs prev", format = colFormat(percent = TRUE, digits = 0)),
      ru_stars      = colDef(name = "Stars", cell = fmt_short),
      ru_dependents = colDef(name = "Used by", minWidth = 80, cell = fmt_short),
      deadline      = colDef(name = "Deadline", minWidth = 100, cell = \(value) if (is.na(value)) "" else span(class = "bad", format(value))),
      archived_n    = colDef(name = "Archived", minWidth = 100, cell = \(value, i) {
        if (is.na(value) || value == 0) return("")
        span(title = coalesce(tbl$cran_comment[i], ""), value, muted("last ", format(tbl$archived_last[i])))
      }),
      needs_compilation = colDef(name = "Compiles", minWidth = 85, cell = \(value) if (isTRUE(value)) "yes" else ""),
      deps_n            = colDef(name = "Deps"),
      deps_compiled_n   = colDef(name = "Compiled", minWidth = 85),
      sysreqs = colDef(name = "System reqs", minWidth = 280, cell = \(value, i) {
        via <- tbl$sysreqs_via_deps[i]
        det <- tbl$sysdeps_detected[i]
        tagList(
          if (!is.na(value)) div(value),
          if (!is.na(via)) muted(title = via, "via deps: ", str_trunc(via, 90)),
          if (!is.na(det)) muted("links: ", det)
        )
      }),
      revdeps_n   = colDef(name = "Direct"),
      revdeps_top = colDef(name = "Top by score", minWidth = 220),
      flags       = colDef(name = "Flags", minWidth = 160, cell = flag_badges),
      notes       = colDef(name = "Notes", minWidth = 200)
    )
  )
)
```

## Movers

```{r}
snaps    <- list.files("../data/snapshots", pattern = "\\.csv$", full.names = TRUE) |> sort()
run_date <- as.Date(run$run_date)
old_path <- snaps |> keep(\(p) as.Date(str_remove(basename(p), "\\.csv$")) <= run_date - 28) |> tail(1)
old      <- if (length(old_path)) read_snapshot(old_path)
```

`r if (is.null(old)) "_Score and download trends show up once there are snapshots 4+ weeks apart._" else glue("Biggest r-universe score changes since {old$run_date[1]}:")`

```{r}
if (!is.null(old)) {
  snap |>
    select(package, ru_score, dl_365) |>
    inner_join(old |> select(package, old_score = ru_score, old_dl = dl_365), by = "package") |>
    mutate(score_change = ru_score - old_score, dl_change = dl_365 / old_dl - 1) |>
    filter(!is.na(score_change)) |>
    slice_max(abs(score_change), n = 10, with_ties = FALSE) |>
    select(package, ru_score, score_change, dl_change) |>
    reactable(compact = TRUE, columns = list(
      ru_score     = colDef(name = "Score", format = colFormat(digits = 1)),
      score_change = colDef(name = "Change", format = colFormat(digits = 2)),
      dl_change    = colDef(name = "Downloads change", format = colFormat(percent = TRUE, digits = 0))
    ))
}
```

## Links

Not R packages, but related. From [links.yml](https://github.com/`r repo`/blob/main/links.yml).

```{r}
#| results: asis
links$links |> map_chr(\(l) glue("- [{l$name}]({l$url}){note(l)}")) |> cat(sep = "\n")
```

### Unresolved

Names from the watchlist I couldn't identify yet:

```{r}
#| results: asis
links$unresolved |> map_chr(\(l) glue("- **{l$name}** ({l$theme}){note(l)}")) |> cat(sep = "\n")
```
