---
title: "Themes"
subtitle: "What each group of packages is for, how they overlap, and what I'd pick"
toc: true
toc-depth: 2
---

```{r}
#| include: false
for (f in list.files("../R", pattern = "\\.R$", full.names = TRUE)) source(f)
snap  <- read_snapshot("../data/latest.csv")
watch <- read_watchlist("../packages.yml")
notes <- read_watch_notes("../packages.yml")
repo  <- Sys.getenv("GITHUB_REPOSITORY", "cgwweaver/cranwatch")
```

Short guides for colleagues, one per theme in [`packages.yml`](https://github.com/`r repo`/blob/main/packages.yml). Each table is live from the latest run (`r snap$run_date[1]`), and notes come from the comments in packages.yml. Click a package for its CRAN or r-universe page. All columns are on the [Radar](index.html), and [How to judge a package](judge.html) explains how to read them.

Guides marked [draft]{.flag} haven't been reviewed by a human yet.

```{r}
#| results: asis
watch |>
  distinct(theme) |>
  pull() |>
  walk(\(th) {
    guide <- read_guide("../guides", th)
    theme_note <- notes |> filter(theme == th, is.na(package)) |> pull(note)

    rows <- watch |>
      filter(theme == th) |>
      select(package) |>
      left_join(snap, by = "package") |>
      left_join(notes |> filter(theme == th, !is.na(package)) |> select(package, note), by = "package") |>
      arrange(desc(ru_score)) |>
      transmute(
        Package  = md_pkg(package, link),
        Status   = status,
        Score    = if_else(is.na(ru_score), "", format(round(ru_score, 1), nsmall = 1)),
        `DLs/yr` = fmt_short(dl_365),
        Deps     = if_else(is.na(deps_n), "", glue("{deps_n} ({deps_compiled_n} compiled)")),
        Flags    = coalesce(flags, ""),
        Note     = coalesce(note, "")
      )

    cat("\n\n## ", th, if (guide$draft) " [draft]{.flag}", "\n\n", sep = "")
    if (length(theme_note)) cat("_", theme_note, "_\n\n", sep = "")
    cat(if (is.null(guide$text)) "_No guide yet._" else guide$text, sep = "\n")
    cat("\n\n")
    print(knitr::kable(rows, format = "pipe"))
    cat(glue("\n\n[Edit this guide](https://github.com/{repo}/edit/main/guides/{theme_slug(th)}.md)\n\n"))
  })
```
